Drought-sensitive trees survive on deep-water access in a tropical forest community

### 2020-09-19 manuscript.rmd
incorporated some of the co-authors' comments

---
title: "Drought-sensitive trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

- Among the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds. Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a new community-scale model of species leaf vulnerability curves. We evaluated ERD against existing isotopic tracer data on trees and soil. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees mitigate water-stress related mortality risk through deep water access.
- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.

@@

# Introduction
\
Tropical forests play a disproportionately large role in the global carbon and energy cycle, and support half of global biodiversity [@Bonan:2008; @Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity in mitigating mortality events. Despite this, widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies below-ground by vertical depth and so do rooting depths of co-occurring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018]. Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses. A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~70-90 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water with depth. However, such data are extremely rare [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.

\
Trade-offs shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010]. This entails identifying trade-offs among environment, traits and demography. Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011]. Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density; see Table 1 for trait symbols and their definitions) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear. Deep rooting may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].

\
Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. While correlations between aboveground hydraulic traits and mortality have been identified [@Anderegg:2016], numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]. However, plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region. To better understand how below-ground rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, we make a thorough assessment for a tropical forest of trade-offs between the below-ground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration. Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum optimized for increasingly reliable water availability by depth. Our hypotheses are that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductivity ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information.

\
In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model. We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for 21 species. Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; see Table \@ref(tab:symbols) for definitions) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of 321 total). We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and tested coordination of ERD with hydraulic traits and mortality rates.

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Although 28% of canopy tree species are at least partly deciduous during the dry season, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

@@

## Leaf Vulnerability Curves
\
We measured ($K_{leaf}$ vs. $\Psi_{leaf}$) on adult trees of `r length(sp.exp.param$sp)` common species (see [SI](#SIData)). We fitted four different functions to species-specific data using the *anneal* optimization function with the package likelihood in R: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$) [@Guyot:2011]. The differences among AICc values for these functions were not large within species (Zailaa et al. to be submitted). We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:

@@

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Pairwise correlations among Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) and hydraulic properties for seven large, canopy species found on Barro Colorado Island (Panama). Scatterplots with linear fits are shown in the lower-left triangle. Corresponding Spearmon's $\\rho$ are given in the diagonally opposite location in the upper-right triangle. *P*-values $\\le$ 0.01, 0.05 and 0.1 are represented by symbols **, * and . , respectively. Diagonal shows density distribution of each variable.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.chart.jpeg"))
```

@@

**RCT: Upload on NGET-Archive and site DOIs:
(1) Boris to upload climate forcing data
(2) Rutuja to upload R scripts for ELM-FATES parameter generations, ELM-FATES modified code, and all R scripts to reproduce manuscript.
\
Barro Colorado Forest Census Plot Data can be obtained online[@Condit.2019a;@Condit.2019b].

@@

### Data at the Eddy Flux Tower
Matteo to add. ET, TDR and may be vary briefly, GPP and VPD.

### $K_{leaf}$ vs. $\Psi_{leaf}$
Joseph et al. to add.

### $K_{stem}$ vs. $\Psi_{stem}$
Brett to add.

@@

```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in Leaf Cover Fraction for ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness != "Evergreen"]))," non-evergreen species at Barro Colorado Island. Leaf Cover Fraction for evergreen species (*n* = ", length(unique(sp.leaf_cover.for.model$sp[sp.leaf_cover.for.model$deciduousness == "Evergreen"])),") is assumed at 1 through the year.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf_fall/leaf.cover_BCI_single_panel_evergreensAt100.jpeg"))
```

@@

```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) depth z over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines). Vertical gray lines show median dates of the demographic censuses at the BCI 50 ha plot."}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
```

@@

>>>>

---
title: "Drought-sensitive trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

- Among the mechanisms that plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied. Vulnerability to embolism and hydraulic safety margins are shown to be predictors of drought-induced mortality risk, but deep-water access may delay plants from reaching these critical thresholds. Here, we tested the role of deep-water access in survival strategies of a diverse tropical forest community (Panama) using a novel data-model approach. 
- We inversely modeled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy species by linking growth dynamics (1990-2015) to vapor pressure deficit, water potentials in the whole-soil column, and a species-specific leaf vulnerability curves. We evaluated ERD against existing isotopic tracer data on trees and soil. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that trees mitigate water-stress related mortality risk through deep water access.
- The role of deep-water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of forest dynamics under current and future climate.

@@

# Introduction
\
Tropical forests play a disproportionately large role in the global carbon and energy cycle [@Bonan:2008], and support half of global biodiversity [@Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are logically considered to be an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw], yet mortality events in the tropics are rarely as large as those in the temperate and boreal zones (McDowell et al. 2018), leading to questions regarding the role of large trait-diversity in mitigating mortality events. Despite this, widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010] and tropical mortality rates are increasing in some regions [@Brienen.2015; @Hubau.2020]. However, state-of-the-art dynamic vegetation models (DVMs) struggle to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013; @Powell:2018], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies below-ground by vertical depth, and that the rooting depths of colocated trees also vary, by species and size [@Meinzer:1999; @Chitra-Tarak.2018]. Consequently, trees in the same forest differ in their drought experience, and thus in their growth and mortality responses. A key bottleneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 60 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~70-90 % (40 - 53 thousand species) of the world’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect measure of water sourcing depths by matching the isotopic value in plant tissue to variations in soil pore water with depth. However, such data are extremely rare [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.

\
Trade-offs in plant 'traits' shape how plants make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010]. This entails identifying relationships among environment, traits and demography. Specifically for the water environment (or, resource), observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to those for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and below-ground traits or environment linkages are weak [@Reich:2014; @Sterck:2011]. Acquisitive species have traits with high capacity to acquire, move, store and use water (higher hydraulic conductivity, low wood density) and achieve high growth, but build flimsy tissues (high SLA) and are less tolerant of low resource availablity, whereas species with conservative traits are better protected from carbon losses and are more physiologically drought tolerant (greater embolsim resistance, larger safety margin). Whether and how the acquisitive-conservative spectrum holds true for the water dynamics vertically below-ground is not clear. Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019]. Deep rooting may be expected to support an acquisitive strategy by enabling a steady water supply. However, the degree to which deep rooting is associated with an acquisitive strategy may depend on the exact dynamics of water availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven systems *shallow*-rooted plants are associated with acquisitive traits (higher hydraulic conductivity) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013 and references therein].

\
Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. While correlations between aboveground hydraulic traits and mortality have been identified [@Anderegg:2016], numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]. However, plants' realized water environments, specifically their ability to acquire water via differing rooting depths, are rarely studied, especially in the tropical region.

\
To better understand how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tropical forest tree species, in this paper, we conduct a thorough assessment for a tropical forest of relationships between the water environment in the whole-soil column, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that witnessed droughts of various intensity and duration. Specifically, we test an overarching hypothesis that inter-specific variation in trees' effective rooting depths aligns species on an acquisitive-conservative economic spectrum, . whereby acquisitive trees have rooting depths optimized for increasingly reliable water availability by depth. Our hypotheses are that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem, whereas shallower ERD species are associated with less efficient but safer hydraulic traits (see Table \@ref(tab:hypo)); and that, (ii) deeper ERD species have lower mortality rates due to low realized risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

\
```{r hypo, results = 'asis', echo = FALSE}
knitr::kable(hypo.table, caption = 'Hypotheses for association between Effective Rooting Depth (ERD) and hydraulic traits.', booktabs = TRUE, escape = FALSE)
```

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Descriptions of datasets used and details of some of the methods are described in the Supplementary Information.

\
In summary, we obtained daily dynamics of plant water potential by depth in the whole-soil column with a locally parameterized and calibrated 1-D hydrological water balance model. We developed a novel inverse model to estimate species effective rooting depths (ERD), defined as the depth at which soil water dynamics best explained observed dynamics of species growth. This model incorporated the impact of atmospheric drought on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves. For the latter, we collected new data on vulnerability curves for 21 species and scale-up to the community via scaling relationships with 'softer' traits, namely, Wood Density and Leaf Mass Area. We evaluated ERD against independent estimates based on existing data on plant and soil water isotopes. We then tested whether deeper ERD was associated with hydraulically efficient but vulnerable traits (Table 1 ), but not greater mortality rates.

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Although 28% of canopy tree species are at least partly deciduous during the dry season, this amounts to the maximum percentage of deciduous individuals in the canopy this amounts to is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

@@

## Leaf Vulnerability Curves
\
We obtained leaf hydraulic vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$; see Table \@ref(tab:symbols) for trait symbols and their definitions) for adult trees of 21 common species from Zailaa et al. (to be submitted; see SI for brief description of methods).

We chose the exponential function without the intercept to fit observed data across species, and as a general function to scale-up to the community, because it had the lowest AICc for most species, captured the form of the observed data across all species well, and had the lowest number of parameters, that is, two:

@@

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for ", nrow(cor.data)," canopy species found on Barro Colorado Island (Panama). Each circle is color coded by the Spearman's $\\rho$ with darker red or blue shades for increasngly positive or negative $\\rho$ values, respectively. P-values are given only when they are $\\ge$ 0.05. Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/erd.stem.traits_cor.matrix_all_p-val.jpeg"))
```

@@

**RCT: Upload on NGET-Archive and site DOIs:
(1) Boris to upload climate forcing data
(2) Rutuja to upload R scripts for ELM-FATES parameter generations, ELM-FATES modified code, and all R scripts to reproduce manuscript.
\
Barro Colorado Forest Census Plot Data can be obtained online[@Condit.2019a; @Condit.2019b].

@@

### Microclimatic tower

The tower used for these measurements is 41 m above ground, on a plateau on BCI [@Pau.2018]. The eddy covariance system includes a sonic anemometer (CSAT3, Campbell Scientific, Logan, UT) and an open-path infrared CO2/H2O gas analyzer (LI7500, LiCOR. Lincoln, NE). Hi-frequency (10Hz) measurements were acquired by a datalogger (CR1000, Campbell Scientific) and stored on a local PC. Data were processed with a custom program using a standard routine described in [@Detto.2010]. QA/QC criteria for removing erroneous values are listed in table 1 (see below). GPP was derived from daytime values of NEE by adding the corresponding mean daily ecosystem respiration obtained as the intercept of the light response curve [@Lasslop.2010]. The light curve was fitted on a 15-days moving window using a rectangular hyperbolic function (runs with friction velocity less than 0.4 m s^-1^ were excluded). In order to compute daily time integrated budgets, gaps were filled using Artificial Neural Network [@Papale.2003] with hydro-meteorological inputs as predictors (soil moisture, solar radiation, temperature, VPD and air pressure, all measured on the same tower). To train the network, the dataset was randomly divided in a training set (70%), a validation set (15%) and a test set (15%). A two-layer feed-forward network with 10 sigmoid hidden neurons and linear output neurons, was trained using the Levenberg-Marquardt algorithm until the mean square error (MSE) of the validation set stop improving [@Hagan.1994]. Performance, in term of MSE, was evaluated using the test set at the end of the training. This procedure was repeated 100 times to produce 100 estimates of GPP. Training multiple times generates different results due to different initial conditions and random sampling of the three sets. Ensemble was obtained as weighted average from the 100 ANN predictions using the MSE of the test set as weights according to:
f_i=(∑_k▒f_ik^ANN /MSE_ik)/(∑_k▒〖1/MSE_ik 〗)

The ANN was implemented using the Neural Network Toolbox in Matlab 2014a.

### Leaf maximum hydraulic conductivity and vulnerability to cavitation

The Kleaf was measured under steady state and high irradiance following the Evaporative Flux Method (FEM) [@Sack:2012]  using a flowmeter. Briefly, the leaf is excised under filtered DI water and attached to a water source (graduated cylinder) using clear plastic tubing. The system includes tubing of a known resistance, separated upstream and downstream by two sensors measuring the transpiration-driven water flow pressure in the EFM system [@Sack.2011]. The K~leaf~ is calculated as the ratio of the transpiration rate (E~leaf~; mmol s^-1^) to the difference in flow pressure (MPa) upstream and downstream from the resistance tubing, and further standardized to 25°C and by leaf area. At stability (CV < 5%), the leaf is left to equilibrate in plastic bags with moist paper towels and measured for \Psi~leaf~ using a Scholander pressure chamber. To construct vulnerability curves for each species, four different functions were fitted to species-specific K~leaf~ vs. \Psi~leaf~ data using the *anneal* optimization function in the R package likelihood: exponential without an intercept ($K_{leaf} = Ae^{-B\Psi_{leaf}}$), exponential with an intercept ($K_{leaf} = C + Ae^{-B\Psi_{leaf}}$), logistic ($K_{leaf} = \frac{A}{1 + \left(\frac{\Psi_{leaf}}{x_{0}}\right)^B}$) and sigmoidal ($K_{leaf} = \frac{A}{1 + e^{-\left(\frac{\Psi_{leaf} - x_{0}}{B}\right)}}$). The differences among AICc values for these functions were not large within species (Zailaa et al. to be submitted; (see Zailaa et al. for more detailed methods description).

### Stem maximum hydraulic conductivity and vulnerability to cavitation

Brett to edit.

Methods for Stem Maximum hydraulic conductivity and vulnerability to cavitation closely followed [Christman.2012] and [Wolfe.2016]. Values for some of the species reported here were previously reported by @Dickman.2019 and @Wu.2020. The entire dataset and R script for the analyses are archived in the NGEE-Tropics archive (Wolfe 2020, To be done).
\
Canopy branches were cut from 4–10 trees of 26 species across three sites (Table SXX). The branch segments were enclosed in opaque plastic bags, and brought to the laboratory for measurements. Stem segments used for hydraulic conductivity (Kh) measurements were located at least 1 m from the cut end of the branch to minimize the number of open vessels that extended into the segment. Maximum vessel length was measured on a subset of samples from each species. Among species, maximum vessel length ranged XXX and length of stem between the cut that was made in the air to collect the branch and the section of stem used for Kh measurement was XXX (Table SXX).
\
Three leaves that were distal to the Kh segment were sealed in plastic bags and the entire branch was enclosed in black plastic bags for at least 30 minutes to stop transpiration. Just prior to measurement of Kh, leaf water potential of the three bagged leaves was measured with a pressure chamber. Since transpiration was stopped by bagging the leaves, leaf water potential was assumed to equal stem water potential (REFERENCE).
\
For each Kh segment, three measurements of stem water potential were made on stem sections that were located adjacent to the Kh segment. Stem sections that were approximately 40 mm in length were placed in nickel-plated aluminum chambers that were connected to Peltier-type psychrometers (Merrill Instruments, Logan, UT) and placed in a water bath at 30°C for 3 h to equilibrate the humidity within the cambers. Then water potential was measured using an automated multi-channel micro-voltmeter (CR7, Campbell Scientific, Logan, UT) with a 10 s thermocouple cooling current. For samples with water potential < -4 MPa, a 45 s cooling current was used to improve accuracy. Each psychrometer probe was calibrated against salt solutions beforehand and the plateau of the psychromatic response curve was used to assess water potential.
\
The Kh segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)). Then the ends were shaved clean with a fresh razor. Segments were 11–34 cm in length (mean ± SD = 22 ± 4 cm) and 0.2–1.6 cm in diameter (0.7 ± 0.2 cm). The segments were placed in a hydraulic conductivity apparatus similar to that described by Sperry et al. (1998) except that flow rate was measured with graduated pipettes instead of a balance. The perfusion solution was 10 mM KCl filtered to 0.2 m. For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure. Following Torres-Ruiz et al. (2012), Kh was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads. Stem area specific hydraulic conductivity was (Ks) was calculated by dividing Kh by the cross-sectional area of stem. To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.
\
To assess vulnerability to cavitation, the branches were allowed to dry in the lab for 0 – 180 hours before they were measured for stem water potential and Ks. We used the water potential measured by the psychrometers because they are a more direct measurement. It has been noted that bagged-leaf water potential can deviate from stem water potential, especially when branch segments are relatively dry (REFERENCE). For some species, we noted this pattern while for others it was not apparent (Fig SXX). For each species, Ks was plotted against stem water potential and a Weibull curve was fit through the 90th  percentile of the points, following Wolfe et al. (2016)). The intercept of the curve was interpreted as maximum Ks and the water potential where the Weibull curve predicted 12, 50, 80, and 88 loss of maximum Ks were interpreted as P12, P50, P80, and P88, respectively.

@@

```{r leaf-cover, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-leaf-cover}Seasonal variation in Leaf Cover Fraction for ", length(erd.sp)," species at Barro Colorado Island (panels). Species are color-coded by deciduousness.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/leaf.cover_BCI_multi_panel.jpeg"))
```

@@

```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi_{soil,~z}$) by depth z (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines). Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
```

@@

```{r eddy, results = 'asis', echo = FALSE}
knitr::kable(eddy.table, caption = 'QA/QC procedure applied to eddy covariance fluxes. w: vertical wind speed; t: air temperature; q: water vapor density; c: CO~2~ density; u: horizontal wind speed. The bar indicates 30-min averaging and the apex indicates fluctuation ($\\overline{x` = x - \\bar{x}}$)', booktabs = TRUE, escape = FALSE, col.names = gsub("[_]", "-", gsub("[.]", " ", names(eddy.table))))
```
